---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 18 @ 11:59PM
author: Yuhang Qian
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Display machine information:
```{r}
sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(miceRanger))
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

    A more thorough book treatment of the practical imputation strategies is the book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.

***MCAR, missing completely at random, the missing data are unrelated to the study variables: thus, the participants with completely observed data are in effect a random sample of all the participants assigned a particular intervention.***

***MAR, missing at random, whether or not data are missing may depend on the values of the observed study variables. However, after conditioning on this information, whether or not data are missing does not depend on the values of the missing data.***

***MNAR, missing not at random, whether or not data are missing depends on the values of the missing data.***


2. Explain in a couple of sentences how the Multiple Imputation by Chained Equations (MICE) work.

***MICE is a multiple imputation method used to replace missing data values in a data set under certain assumptions about the data missingness mechanism (e.g. MAR). Consider an imputation strategy that moves from variable to variable, treating each as a separate missing data problem: Specify a prediction equation of an appropriate form (e.g. linear, binary, etc) for each variable in turn; Randomly draw from the conditional distribution specified by the regression model; Replace previously imputed values with new draws; Iterate for a predetermined number of cycles.***



3. Perform a data quality check of the ICU stays data. Discard variables with substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by `NA`s.

```{r}
# Load the dataset
icu_cohort <- readRDS("../hw3/mimiciv_shiny/icu_cohort.rds")
summary(icu_cohort)
```



4. Impute missing values by `miceRanger` (request $m=3$ data sets). This step is computational intensive. Make sure to save the imputation results as a file. Hint: Setting `max.depth=10` in the `miceRanger` function may cut some computing time.

5. Make imputation diagnostic plots and explain what they mean.

6. Choose one of the imputed data sets to be used in Q2. This is **not** a good idea to use just one imputed data set or to average multiple imputed data sets. Explain in a couple of sentences what the correct Multiple Imputation strategy is.

## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression (`glm()` function in base R or keras), (2) logistic regression with lasso penalty (glmnet or keras package), (3) random forest (randomForest package), or (4) neural network (keras package).

1. Partition data into 80% training set and 20% test set. Stratify partitioning according the 30-day mortality status.

2. Train the models using the training set.

3. Compare model prediction performance on the test set.

